---
title: "pharmetrix_eda"
output: html_document
date: "2024-10-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## import packages

```{r}
library(ohdsilab)
library(tidyverse)
library(DatabaseConnector)
library(keyring)
library(FeatureExtraction)
library(stats)
library(purrr)
library(table1)
library(easystats)
library(glue)
```

## Set up connection to database

```{r}
cdm_schema = "omop_cdm_53_pmtx_202203"
write_schema = paste0("work_", keyring::key_get("db_username"))

# Create the connection
con =  DatabaseConnector::connect(
  dbms = "redshift",
  server = "ohdsi-lab-redshift-cluster-prod.clsyktjhufn7.us-east-1.redshift.amazonaws.com/ohdsi_lab",
  pathToDriver = 'D:/Users/tilton.ca',
  port = 5439,
  user = keyring::key_get("db_username"),
  password = keyring::key_get("db_password")
)

# Test if the connection words
if(isTRUE(DatabaseConnector::dbIsValid(con))){print("Connected Successfully")}

# make it easier for some r functions to find the database
options(con.default.value = con)
options(schema.default.value = cdm_schema)
options(write_schema.default.value = write_schema)
```
## set schema and table names for the sql queries

```{r}
personal_schema <- "work_tilton_ca204"
model_table_name <- "full_model_1_demo"
insurance_table_name <- "model_1_insurance_demo"
cohort_table_name <- "stroke_cohort_w_conditions_demo"
```


## Use getdbcovariateData() to calculate Charlson Comorbidity index for all patients in the cohort

```{r}
#  settings <- createCovariateSettings(
#    useCharlsonIndex=TRUE
#  )
# 
# # Create the connection
# 
# connectionDetails<-createConnectionDetails(
#  dbms="redshift",
#  server= "ohdsi-lab-redshift-cluster-prod.clsyktjhufn7.us-east-1.redshift.amazonaws.com/ohdsi_lab",
#  pathToDriver = 'D:/Users/tilton.ca',
#  port = 5439,
#  user = keyring::key_get("db_username"),
#  password = keyring::key_get("db_password")
#  )
# 
# 
#  covariateData <- getDbCovariateData(
#  connectionDetails = connectionDetails,
#  cdmDatabaseSchema = "omop_cdm_53_pmtx_202203",
#  cohortDatabaseSchema = "work_tilton_ca204",
#  cohortTable = "stroke_cohort_w_conditions_test",
#  rowIdField = "subject_id",
#  cohortIds = -1,
#  covariateSettings = settings
#  )
#  summary(covariateData)

```


```{r}
#covariateData$covariates
```

# save the CCI data in a file
```{r}
#saveCovariateData(covariateData, "covariates")
```

## load the CCI data file
```{r}
#covariate_file <- loadCovariateData("covariates")
covariate_file <- loadCovariateData("D:/Users/tilton.ca/Documents/r-ohdsi/Casey_Tilton_OHDSI/covariates")
CCI_df <- covariate_file$covariates
CCI_df
```

## plotting the distribution of CCI. 
Due to the right tail, I will use median to impute missing values, not mean
```{r}
ggplot(CCI_df, aes(x = covariateValue) ) + 
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Charlson comorbidity index", x = "CCI", y = "Count of patients") +
  theme_minimal()

```

```{r}
ggplot(CCI_df, aes(x = "", y = covariateValue)) +
  geom_boxplot(fill = "steelblue", color = "black") +
  labs(title = "Boxplot of Numbers", x = "", y = "Numbers") +
  theme_minimal()
```



## function that calculates and plots confidence intervals for model parameters

```{r}
age_labels <- c("18-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80+")
plot_model_coefficients <- function(model, age_group_labels, model_name) {
  # Extract coefficients and confidence intervals
  coefficients <- summary(model)$coefficients
  confint_vals <- confint(model)
  
  # Prepare data for plotting
  plot_data <- data.frame(
    Group = rownames(coefficients)[-1], # Remove the intercept row
    Coefficient = coefficients[-1, 1], # Extract coefficients excluding intercept
    LowerCI = confint_vals[-1, 1],     # Exclude intercept from confint
    UpperCI = confint_vals[-1, 2],     # Exclude intercept from confint
    Significance = coefficients[-1, 4] < 0.05 # Exclude intercept significance
  )
  
  # Identify AGE_GROUP contrasts and relabel them
  age_group_rows <- grep("^AGE_GROUP", plot_data$Group)
  num_groups <- length(age_group_labels) - 1 # Sum contrast has one less group
  if (length(age_group_rows) != num_groups) {
    stop("Number of AGE_GROUP contrasts does not match the number of age group labels minus one.")
  }
  plot_data$Group[age_group_rows] <- paste0(age_group_labels[1:num_groups], " vs AGE_GROUP grand mean")
  
  # Create the plot
  plot <- ggplot(plot_data, aes(x = Group, y = Coefficient)) +
    geom_point(aes(color = Significance), size = 3) +
    geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI, color = Significance), width = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.8) +
    scale_y_continuous(breaks = seq(floor(min(plot_data$LowerCI)), 
                                    ceiling(max(plot_data$UpperCI)), 
                                    by = 0.25)) + # Set ticks every 0.25
    theme_minimal() +
    labs(title = "Confidence Intervals for Parameters in Model",
         x = "Model Parameters",
         y = "Coefficient") +
    coord_flip() +
    scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "orange"), 
                       name = "Significant at <.05") +
    theme(axis.text.x = element_text(size = 14),
          axis.text.y = element_text(size = 14),
          plot.title = element_text(size = 14, face = "bold"))
  
  # Display the plot
  print(plot)

  # Save the plot as a PNG file
  file_path <- paste0("D:/Users/tilton.ca/Documents/github/project-cbt87/figs/", model_name, ".png")
  ggsave(filename = file_path, plot = plot, width = 10, height = 7, dpi = 300)
}
```

## crosstabs function
```{r}
# List of predictors
predictors <- c("HAS_PARKINSONS", "HAS_MIGRAINE", "HAS_DEPRESSION","HAS_INSOMNIA", "HAS_DEMENTIA", "HAS_PARALYSIS","HAS_ISCHEMIC_STROKE", "SEVERITY", "GENDER", "AGE_GROUP")

# Crosstab function
generate_crosstab <- function(predictor, df, target_variable) {
  crosstab <- table(df[[predictor]], df[[target_variable]])
  chisq_result <- chisq.test(crosstab) # Chi-squared test
  list(
    crosstab = crosstab,
    chisq_p_value = chisq_result$p.value
  )
}
```

## import previously created model dataset from personal schema

```{r}

sql_query <- glue("select * from {personal_schema}.{model_table_name}")
full_model_1 <- querySql(con, sql_query)
```

```{r}
full_model_1
```

## import previously created insurance plan data for each model 1 cohort from personal schema

```{r}
sql_query <- glue("select * from {personal_schema}.{insurance_table_name};")
insurance_table <- querySql(con, sql_query)
insurance_table
```



## left join the model dataframe created in dbeaver with the Charlson comorbidity index data, There are 1623 patients (~6.3%) who didn't have a CCI index(not sure why). Next, inner join the insurance plan data by person_id. This pipeline removes the unnecessary covariateid column, renames the covariatevalue column to CCI, turns the target variable into binary, imputes the median CCI value into the rows that are missing data, removes patients under 18, and breaks the patient_age variable into age bins. Then it codes the AGE_GROUP variable using sum-to-zero contrast coding to compare each group as how it deviates from the grand mean. 

```{r}
age_breaks <- c(18, 30, 40, 50, 60, 70, 80, Inf)
age_labels <- c("18-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80+")


model_1_df <- left_join(full_model_1, df_real, by = c("SUBJECT_ID" = "rowId")) |>
  inner_join(insurance_table, by = "SUBJECT_ID") |>
  select(-covariateId) |>
  rename(CCI = covariateValue) |>
  mutate(SLP_TREATMENT_STATUS = ifelse(SLP_TREATMENT_STATUS == "SLP_treatment", 1,0)) |>
  mutate(CCI = replace_na(CCI, median(CCI, na.rm = TRUE))) |>
  filter(PATIENT_AGE >= 18) |>
  mutate(
    AGE_GROUP = cut(
      PATIENT_AGE,
      breaks = age_breaks,
      labels = age_labels,
      right = FALSE,
      ordered_result = FALSE))
  


contrasts(model_1_df$AGE_GROUP) <- contr.sum(length(levels(model_1_df$AGE_GROUP)))


model_1_df
```

# Creating the Descriptive Statistics Table (aka Table 1)

## bring in the full cohort table

```{r}
sql_query <- glue("select * from {personal_schema}.{cohort_table_name};")
cohort_with_conditions <- querySql(con, sql_query)
cohort_with_conditions
```


## Using the pivot_longer() function to create a row for each condition per person

```{r}

cols_to_convert <- c("SLP_TREATMENT_STATUS", "HAS_PARKINSONS", "HAS_MIGRAINE", "HAS_DEPRESSION", "HAS_INSOMNIA", "HAS_DEMENTIA", "HAS_PARALYSIS", "HAS_ISCHEMIC_STROKE")

model_1_df_longer <- inner_join(cohort_with_conditions, model_1_df, by = "SUBJECT_ID") |>
  pivot_longer(
    cols = c(HAS_APHASIA, HAS_DYSPHAGIA, HAS_DYSARTHRIA), # Specify exact column names
    names_to = "SPEECH_CONDITION", # Name for new column
    values_to = "HAS_CONDITION" # Name for values
  ) |>
  filter(HAS_CONDITION == 1) %>% # Keep rows where the condition is present
  mutate(SPEECH_CONDITION = str_remove(SPEECH_CONDITION, "HAS_") %>% # Remove prefix
           tolower()) |> # Convert to lowercase
  mutate(across(all_of(cols_to_convert), as.factor))

model_1_df_longer
```

## generate table1

```{r}
overall_df <- model_1_df_longer |>
  distinct(SUBJECT_ID, HAS_PARKINSONS, HAS_MIGRAINE, HAS_DEPRESSION, HAS_INSOMNIA, HAS_DEMENTIA, HAS_PARALYSIS, HAS_ISCHEMIC_STROKE, SEVERITY, GENDER, AGE_GROUP, PATIENT_AGE, CCI, INSURANCE_PLAN, SLP_TREATMENT_STATUS) |>
  mutate(SPEECH_CONDITION = "Total")

# Combine "Overall" group with original data
df_combined <- bind_rows(model_1_df_longer, overall_df)

# Generate the table
label(df_combined$HAS_PARKINSONS) <- "Parkinson's Disease"
label(df_combined$HAS_MIGRAINE) <- "Migraine"
label(df_combined$HAS_DEPRESSION) <- "Depression"
label(df_combined$HAS_INSOMNIA) <- "Insomnia"
label(df_combined$HAS_DEMENTIA) <- "Dementia"
label(df_combined$HAS_PARALYSIS) <- "Paralysis"
label(df_combined$HAS_ISCHEMIC_STROKE) <- "Ischemic Stroke"
label(df_combined$SEVERITY) <- "Severity"
label(df_combined$GENDER) <- "Gender"
label(df_combined$AGE_GROUP) <- "Age Group"
label(df_combined$PATIENT_AGE) <- "Age *"
label(df_combined$CCI) <- "Charlson Comorbidity Index"
label(df_combined$INSURANCE_PLAN) <- "Insurance Plan"
label(df_combined$SLP_TREATMENT_STATUS) <- "SLP Treatment Status"

footnote <- "* We drop age as a continuous variable in favor of age groups in the final models because patients born before or during 1937 are all binned into one year"

table1(~HAS_PARKINSONS + HAS_MIGRAINE + HAS_DEPRESSION + HAS_INSOMNIA + HAS_DEMENTIA + HAS_PARALYSIS + HAS_ISCHEMIC_STROKE + SEVERITY + GENDER + AGE_GROUP + PATIENT_AGE + CCI + INSURANCE_PLAN + SLP_TREATMENT_STATUS | SPEECH_CONDITION, data = df_combined, overall = FALSE, footnote = footnote)
```

## Splitting the table 1 into two columns so I can display it in a powerpoint slide

```{r}
# Split variables into two groups
table1_part1 <- table1(~ HAS_PARKINSONS + HAS_MIGRAINE + HAS_DEPRESSION + 
                        HAS_INSOMNIA + HAS_DEMENTIA + HAS_PARALYSIS + HAS_ISCHEMIC_STROKE + SEVERITY + GENDER | SPEECH_CONDITION, 
                        data = df_combined, overall = FALSE)

table1_part2 <- table1(~ AGE_GROUP + 
                        PATIENT_AGE + CCI + INSURANCE_PLAN + SLP_TREATMENT_STATUS | SPEECH_CONDITION, 
                        data = df_combined, overall = FALSE, footnote = footnote)

# Save these separately or display side-by-side on the slide.

```

```{r}
table1_part1
table1_part2
```



## model including any patient that has one or more of the three conditions

```{r}
model_any_condition <- glm(SLP_TREATMENT_STATUS ~ HAS_PARKINSONS + HAS_MIGRAINE + HAS_DEPRESSION + HAS_INSOMNIA + HAS_DEMENTIA + HAS_PARALYSIS + HAS_ISCHEMIC_STROKE + SEVERITY + GENDER + AGE_GROUP + CCI, data = model_1_df, family = "binomial")
print(summary(model_any_condition), show.residuals = TRUE)
```

## use the function from above to plot the confidence intervals for the coefficients

```{r}

plot_model_coefficients(model_any_condition, age_labels, "model_any_condition")
```
## Output model results in a formatted table
```{r}
parameters::model_parameters(model_any_condition) |> 
  insight::format_table() |> 
  insight::print_html()
```


## Join the cohort table with the model 1 table

```{r}
model_1_df_joined <-inner_join(cohort_with_conditions, model_1_df, by = "SUBJECT_ID")
model_1_df_joined
```



## filter dataframes to only include patients that have a specific speech condition
```{r}
model_1_df_aphasia <- model_1_df_joined |>
  filter(HAS_APHASIA == 1)
model_1_df_aphasia

model_1_df_dysphagia <- model_1_df_joined |>
  filter(HAS_DYSPHAGIA == 1)
model_1_df_dysphagia

model_1_df_dysarthria <- model_1_df_joined |>
  filter(HAS_DYSARTHRIA == 1)
model_1_df_dysarthria
```


# logistic regression model for aphasia
```{r}
model_aphasia <- glm(SLP_TREATMENT_STATUS ~ HAS_PARKINSONS + HAS_MIGRAINE + HAS_DEPRESSION + HAS_INSOMNIA + HAS_DEMENTIA + HAS_PARALYSIS + HAS_ISCHEMIC_STROKE + SEVERITY + GENDER + AGE_GROUP + CCI, data = model_1_df_aphasia, family = "binomial")
print(summary(model_aphasia), show.residuals = TRUE)
```


```{r}
plot_model_coefficients(model_aphasia, age_labels, "aphasia")
```

```{r}
parameters::model_parameters(model_aphasia) |> 
  insight::format_table() |> 
  insight::print_html()
```


# logistic regression model for dysphagia

```{r}
model_dysphagia <- glm(SLP_TREATMENT_STATUS ~ HAS_PARKINSONS + HAS_MIGRAINE + HAS_DEPRESSION + HAS_INSOMNIA + HAS_DEMENTIA + HAS_PARALYSIS + HAS_ISCHEMIC_STROKE + SEVERITY + GENDER + AGE_GROUP + CCI, data = model_1_df_dysphagia, family = "binomial")
print(summary(model_dysphagia), show.residuals = TRUE)
```

```{r}
plot_model_coefficients(model_dysphagia, age_labels, "dysphagia")
```
```{r}
parameters::model_parameters(model_dysphagia) |> 
  insight::format_table() |> 
  insight::print_html()
```


# logistic regression model for dysarthria

```{r}
model_dysarthria <- glm(SLP_TREATMENT_STATUS ~ HAS_PARKINSONS + HAS_MIGRAINE + HAS_DEPRESSION + HAS_INSOMNIA + HAS_DEMENTIA + HAS_PARALYSIS + HAS_ISCHEMIC_STROKE + SEVERITY + GENDER + AGE_GROUP + CCI, data = model_1_df_dysarthria, family = "binomial")
print(summary(model_dysarthria), show.residuals = TRUE)
```

```{r}
plot_model_coefficients(model_dysarthria, age_labels, "dysarthria")
```
```{r}
parameters::model_parameters(model_dysarthria) |> 
  insight::format_table() |> 
  insight::print_html()
```


## Distribution of Age variable

```{r}
ggplot(model_1_df, aes(x = PATIENT_AGE)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Age Distribution of Model 1 cohort",
       x = "Age",
       y = "Count")+
  theme_minimal()
```
## Distribution of Age Group variable
```{r}


# Define the age breaks and labels
age_breaks <- c(18, 30, 40, 50, 60, 70, 80, Inf)
age_labels <- c("18-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80+")

# Use dplyr to create AGE_GROUP
model_1_df_test <- model_1_df %>%
  mutate(
    AGE_GROUP = cut(
      PATIENT_AGE,
      breaks = age_breaks,
      labels = age_labels,
      right = FALSE,
      ordered_result = TRUE
    )
  )

# View the resulting dataframe
glimpse(model_1_df_test)

ggplot(model_1_df_test, aes(x = AGE_GROUP)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(
    title = "Count of Rows by Age Group",
    x = "Age Group",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for readability
  )

```
## Plot the venn diagram of the breakdown of all model 1 patients and which of the speech conditions they have

```{r}
# Install and load VennDiagram package if not already installed
if (!require(VennDiagram)) {
  install.packages("VennDiagram")
  library(VennDiagram)
}



# Create sets for each condition
aphasia_set <- model_1_df_joined$SUBJECT_ID[model_1_df_joined$HAS_APHASIA == 1]
dysphagia_set <- model_1_df_joined$SUBJECT_ID[model_1_df_joined$HAS_DYSPHAGIA == 1]
dysarthria_set <- model_1_df_joined$SUBJECT_ID[model_1_df_joined$HAS_DYSARTHRIA == 1]

# Calculate intersections
venn_counts <- calculate.overlap(
  list(
    Aphasia = aphasia_set,
    Dysphagia = dysphagia_set,
    Dysarthria = dysarthria_set
  )
)

# Draw the Venn diagram
venn_plot <- draw.triple.venn(
  area1 = length(aphasia_set),
  area2 = length(dysphagia_set),
  area3 = length(dysarthria_set),
  n12 = length(intersect(aphasia_set, dysphagia_set)),
  n23 = length(intersect(dysphagia_set, dysarthria_set)),
  n13 = length(intersect(aphasia_set, dysarthria_set)),
  n123 = length(Reduce(intersect, list(aphasia_set, dysphagia_set, dysarthria_set))),
  category = c("Aphasia", "Dysphagia", "Dysarthria"),
  fill = c("red", "green", "blue"),
  lty = "dashed",
  cex = 2,
  cat.cex = 1.5,
  cat.pos = c(-20, 20, 180)
)

# Save the plot if desired
png("D:/Users/tilton.ca/Documents/github/project-cbt87/figs/venn_diagram.png", width = 800, height = 800)
grid.draw(venn_plot)
dev.off()

```





